---
title: "lab_02"
author: "Brendan Weissel"
date: "2024-09-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse. If you have not installed the tidyverse already, remove the # from the next line and run it first.  
# install.packages('tidyverse')
library(tidyverse)
library(lubridate)
```

## Load Data

For this lab we'll be working with a slice of Maryland voter registration data: records of people who have registration dates this year through Aug. 7. Let's load it.

```{r}
md_voters_2024 <- read_rds("data/lab_02.rds")
```

## Examine Data

Take a look at the data, getting familiar with what it contains. Most of the columns should be self-explanatory, but there are two date columns: STATE_REGISTRATION_DATE and COUNTY_REGISTRATION_DATE. The former represents the date the person registered as a Maryland voter; the latter is when they registered in their current county. They may be the same, but if a voter moves from one county to another they will differ. This is important because if the two dates are the same, that's a pretty good indication that this is a *new* voter, someone who has no history of registering in Maryland. You'll need to keep that definition in mind.

## Answer questions

**Q1** How many voters appear to be *new* registrants since 2024-07-21, when President Joe Biden announced he would leave the race, based on the information above? Create a dataframe to hold the only those voters. Which counties have seen the most? Take a look at county population figures here and compare the two in your answer: https://worldpopulationreview.com/us-counties/maryland.

**A1** There are 17083 new registrants since 2024-07-21. Montgomery county, Baltimore City, Baltimore County and Prince George's county have the most registrants. However this makes sense because these are counties with larger populations in Maryland. The data is pretty much aligned with the population, I did not notice any suprises. 
```{r}
maryland_voter_after_july_21 <- md_voters_2024 |>
 filter(STATE_REGISTRATION_DATE > "2024-07-21" ) |> 
 arrange(STATE_REGISTRATION_DATE)
```

```{r}
new_maryland_voters_by_county <- maryland_voter_after_july_21 |> 
   group_by(COUNTY) |> 
  summarize(voter_count = n()) |> 
  arrange(desc(voter_count))
```


**Q2** A better question is whether the new voters you found in Q1 are an outlier compared to other weeks this year. Using the original dataframe with 2024 registrants, use `mutate()` and lubridate to produce week numbers from the COUNTY_REGISTRATION_DATE (Lubridate has a function that turns a date into a week number. Use that to answer this question). Save that to the original dataframe. Then calculate how many voters are in each week. Describe the distribution of voters groups in a sentence - are the weeks from July 21-August 7 different from the rest, and if so, how?

**A2** The week with the most voters came in June. The second and third most popular weeks was from July 21 - August 7. The top 6 weeks all came in the summer months. The least popular week was in the end of May with only 55 new registrants. Week 28 when Trump was shot was the 4th most popular week. 

```{r}

md_voters_2024 <- md_voters_2024 |>
  mutate(week_number = week(COUNTY_REGISTRATION_DATE))

voters_by_week <- md_voters_2024 |> 
    group_by(week_number) |> 
  summarize(voters_count = n()) |> 
  arrange(desc(voters_count))

```

**Q3** Using lubridate, show which month saw the largest number of registrations. Does the answer help shed more light on Q2?
**A3** Yes this continues to show that the summer months were the most popular. The two events of Trump being shot and Biden dropping out increased the number of registered voters. It would be interesting to see how this compares to previous presidential election years during the summer and if 2024 is an outlier. Oddly once again May is a slow month for registration. 

```{r}
voters_by_month <- md_voters_2024 |> 
    mutate(month_number = month(COUNTY_REGISTRATION_DATE)) |> 
    group_by(month_number) |> 
  summarize(voters_count = n()) |> 
  arrange(desc(voters_count))

```

**Q4** Maryland has, officially, five regions: https://www.visitmaryland.org/article/maryland-regions. Modify the original dataframe again to add a `region` column based on the `COUNTY` value using case_when. Are you going to have to do this for every Maryland county plus Baltimore City? Yes, you are. Which regions and parties account for the most new voters?

**A4** Democrats in Central Maryland and the Democrats Capitol region had the most new voters and Western Maryland had the fewest new voters. The democrats and unnafilliated accounted for the most new voters.  

```{r}
md_voters_2024 <- md_voters_2024 |> 
  mutate(region = case_when(
        COUNTY %in% c("Garrett", "Allegany", "Washington") ~ "Western Maryland",
        COUNTY %in% c("Frederick", "Montgomery", "Prince George's") ~ "Capital Region",
        COUNTY  %in% c("Anne Arundel", "Baltimore City", "Baltimore County","Carroll", "Harford", "Howard") ~ "Central Maryland",
        COUNTY %in% c("Calvert", "Charles", "Saint Mary's")  ~ "Southern Maryland",
        COUNTY %in% c("Kent", "Queen Anne's", "Talbot", "Caroline", "Dorchester", "Wicomico", "Somerset", "Worcester", "Cecil")  ~ "Eastern Shore Maryland"))

voters_by_region_party <- md_voters_2024 |> 
    group_by(region, PARTY) |> 
  summarize(voters_count = n()) |> 
  arrange(desc(voters_count))


all_maryland_voters_by_county <- md_voters_2024 |> 
   group_by(COUNTY) |> 
  summarize(voter_count = n()) |> 
  arrange(desc(voter_count))

all_maryland_voters_by_party <- md_voters_2024 |> 
   group_by(PARTY) |> 
  summarize(voter_count = n()) |> 
  arrange(desc(voter_count))


```

**Q5** What's an additional question you could ask of this data?

**A5** One thing you could look at is if more Women or Men registered to vote. Another thing is to get more specific and see what zip codes had the most new voters. 
