---
title: "Class Reference"
author: "Brendan Weissel"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example, this sets the working directory to a directory called data_journalism_2024_fall inside a directory called jour472 located inside your home directory (the ~ is equivalent of /Users/[youruser] on a Mac).

```{r}
setwd("~/jour472/data_journalism_2024_fall")
```

### How to install and load the tidyverse

```{r}
install.packages("tidyverse")
library(tidyverse)
```
Here is how to add: 
```{r}
2+2
```

Here is how to add a data set from a website:
```{r}
maryland_winred_contributions <- read_rds("https://thescoop.org/files/maryland_winred.rds")
```
Here is how the see the first 6 rows of data: 
```{r}
head(maryland_winred_contributions)
```
Remove duplicate rows
```{r}
 distinct()


```

tr_tot_title() to standarize capitalization in the "city" field

