---
title: "lab_03"
author: "Brendan Weissel"
date: "2024-09-18"
output: html_document
---

```{r setup, include=FALSE}
#knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## Load data

```{r}
# Load MD slice of FEC House candidate contribution data
md_contribs <- read_rds("data/lab_03.rds")
```

## Fix header issues

You'll need to fix one issue with the column names using janitor. Write a code block to do that, making sure that you save the results to the same dataframe.

```{r}
cleaned_contribs <- md_contribs |>
  clean_names()
```


## Answer questions

Q1: Which seven cities in Maryland had the largest *number* of contributions from donors? How many did each have? To answer this question, you'll have to clean the messy "contributor_city" field.

Note: To answer this question, you'll need to see every variation of the "contributor_city" for the top seven cities, which you should identify through some initial analysis. Then find variations of them and change those variations into the most common version.

A1:
BETHESDA	1177			
CHEVY CHASE	705			
BALTIMORE	692			
ANNAPOLIS	492			
POTOMAC	491			
SILVER SPRING	408			
ROCKVILLE	358			


```{r}
Top_seven_city <- cleaned_contribs |> 
group_by(contributor_city) |> 
  summarise(total_contribtions = n()) |> 
  arrange(desc(total_contribtions))

cleaned_contribs <- cleaned_contribs |> 
mutate(contributor_city = case_when(
contributor_city %in% c("BALRIMORE", "BALTMORE", "BLATIMORE") ~ "BALTIMORE",
contributor_city %in% c("SILVER SPRINGS") ~ "SILVER SPRING",
contributor_city %in% c("ANAPOLIS") ~ "ANNAPOLIS",
contributor_city %in% c("POTAMIC") ~ "POTOMAC",
contributor_city %in% c("4615 NORTH PARK AVE APT. 1410") ~ "CHEVY CHASE",

    TRUE ~ contributor_city
))

print(Top_seven_city[1:7, ])


```

Q2: Find the 5-digit ZIP code in Maryland that had the highest total amount of contributions (by dollar value). You will need to use mutate to ensure that you have a zip code column with the first five characters from the existing zip code column.

Where is that zip code and what kind of area is it, demographically? Which committee is the largest recipient of money from that zip code? How many of the top five recipients are Maryland candidates?

A2: 20817. It is a wealthy suburb that includes Bethesda and Potomac. It is mostly middle aged white adults. HELMER FOR CONGRESS is the committee with the largest amount of donations from that Zip. Jamie Raskin and April Delaney are the only Maryland candidates in the top 5 recipients from the top zip.

```{r}
cleaned_zip_codes <- cleaned_contribs |>
  mutate(contributor_zip = str_sub(contributor_zip, start=1L, end=5L))

glimpse(cleaned_zip_codes)

zip_codes_by_contribs <- cleaned_zip_codes |> 
  group_by(contributor_zip) |> 
  summarise(amount_received = sum(contribution_receipt_amount)) |> 
  arrange(desc(amount_received))

top_zip <- 20817

top_committees <- cleaned_zip_codes |> 
  filter(contributor_zip == top_zip) |> 
  group_by(committee_name) |> 
  summarise(amount_received = sum(contribution_receipt_amount)) |> 
      arrange(desc(amount_received))

  print(top_committees[1:5, ])


 
```

Q3: Let's look at top individual donors. Write code to generate a list of top donors using name and city, adding up the total amount of contributions. What does the original data say about the top two donors - who are they and what do they do? Do you see anything in the list of top 10 donors that would make you uncertain about the accuracy of these results?

A3:The top two donors are Melanie Jacobsen and Mark P. Jacobsen respectively. The top two appear to be married. Melanie is a farmer Mark works for Intrafi. Mark is also listed at #5 but without his middle name and William Pugh and Bill Pugh are listed as different people Melanie is also listed sometimes with Mrs. and sometimes without. These issues and others defiently make me uncertain about the accuracy of these results. 

```{r}
top_donors <- cleaned_zip_codes |> 
group_by(contributor_name, contributor_city) |> 
  summarise(total_contributions = sum(contribution_receipt_amount)) |> 
arrange(desc(total_contributions))

print(top_donors[1:5, ])
```
